Thermal reactionomes reveal divergent responses to thermal extremes in warm

### 2016-01-15 ApTranscriptome_TR.Rmd
revisions for BMC Genomics

The raw Illumina fastq files are available from the NCBI short read archive [link tbd]. The assembled transcriptome, annotation and expression values are available from the Harvard Forest Archives, HF-113 [http://harvardforest.fas.harvard.edu/data-archive] datasets [hf113-36](http://harvardforest.fas.harvard.edu/data/p11/hf113/hf113-36-ap-ann-table.csv), [hf113-40](http://harvardforest.fas.harvard.edu/data/p11/hf113/hf113-40-ap-trans.tar) and [hf113-41](http://harvardforest.fas.harvard.edu/data/p11/hf113/hf113-41-ap-gene-exp-quant.tgz). These processed files are used for this analysis due to the computational demands of re-generating the files, but the exact commands for each of these steps are documented below.

>>>>

The raw Illumina fastq files are available from the NCBI short read archive [link tbd]. The assembled transcriptome, annotation and expression values are available from the Harvard Forest Archives, HF-113 [http://harvardforest.fas.harvard.edu/data-archive] datasets [hf113-38](http://harvardforest.fas.harvard.edu/data/p11/hf113/hf113-38-ap-ann-table.txt), [hf113-41](http://harvardforest.fas.harvard.edu/data/p11/hf113/hf113-41-ap-trans.tar) and [hf113-42](http://harvardforest.fas.harvard.edu/data/p11/hf113/hf113-42-ap-gene-exp-quant.tgz). These processed files are used for this analysis due to the computational demands of re-generating the files, but the exact commands for each of these steps are documented below.

### 2016-01-15 doc/ApTranscriptome_manuscript.Rmd
revisions for BMC Genomics

---
title: Thermal reactionomes reveal divergent responses to thermal extremes in warm
  and cool-climate ant species
author: ' '
csl: bmcgenomics.csl
output:
  word_document: default
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: no
    toc_depth: 2
  html_document:
    keep_md: yes
includes: header.tex
bibliography: ApTranscriptome.bib
---

@@

In this study, we used the reactionome method to identify genes that are thermally responsive in two closely-related eastern North American ant species, *Aphaenogaster carolinensis* [@demarco2015] and *A. picea* [@umphrey1996]. *Aphaenogaster* are some of the most common ants in eastern North America [@king2013] where they are keystone seed dispersers [@ness2009; @zelikova2011; @rodriguez-cabal2012]. Ants, and ectotherms in general, have little or no thermal safety margin [@sunday2014] and thus are highly susceptible to climate change [@diamond2011; @kingsolver2013], putting at risk important ecosystem services [@toro2012]. Growth chamber studies have demonstrated that reproduction of *Aphaenogaster* will be compromised by increased tempreatures [@pelini2012], while field studies [@diamond2012] and simulations [@bewick2014] indicate that ant species persistence will depend on combinations of physiology and species interactions. *Aphaenogaster carolinensis* experiences a higher mean annual temperature (MAT) (14.6°C) and less seasonal temperature variation (temperature seasonality = 7,678°) than does *A. picea* (MAT = 4.6°C, seasonality = 10,008°; [@hijmans2005]) at their respective collection sites. In controlled laboratory experiments, these warm- and cold-climate species exhibit corresponding differences in their critical maximum (44.7°C for *A. carolinensis* versus 41.3°C for *A. picea*; see Methods) and minimum temperatures (6.1°C for *A. carolinensis* versus -0.1°C for *A. picea*). These differences between species in their thermal environments and physiological tolerances allowed us to investigate adaptation to both lower and upper thermal extremes in this system.

@@

To summarise, in this project we generated the transcriptomes of two closely-related temperate ant species, and quantified their gene expression across a wide range of thermal conditions. We used a reactionome approach to identify all transcripts that changed expression across a thermal gradient each species. We then evaluated three non-mutually exclusive hypotheses (enhanced response, tolerance and genetic assimilation) of the evolution of thermal adaptation by comparing differences between the thermal reactionomes of each species. Finally, we used gene ontology information to determine which gene products and pathways are involved in thermal adaptation in the two species.

@@

The combined *Aphaenogaster* transcriptome assembly contained `r prettyNum(nrow(annotation.table), big.mark =",")` transcripts. About half of these (`r prettyNum(nrow(annotation.table.responsive), big.mark =",")`) transcripts had a signficant BLAST hit, of which 50% (25,797) had a top hit to Insecta and 37% (18,854) had a top hit to Formicidae. We performed a BUSCO analysis [@simao2015] to assess the quality of the transcriptome assembly against the arthropod Benchmarking Universal Single-Copy Orthologs (BUSCOs). This analysis revealed that transcriptome is largely complete, as we recovered 1,426 complete single-copy BUSCOs (62%) and an additional 435 fragmented BUSCOs (16%), which is in line with results of Simao et al. [-@simao2015] for transcriptomes of other non-model species. Moreover, only 8% of the BUSCOs were found to be duplicated in the transcriptome, which indicates that while we assembled the transcriptomes of the two species together, the steps we took to collapse homologs were successful.

We quantified gene expression using the program `Sailfish` [@patro2014], and fit polynomial regression models to the expression values of each transcript to identify those that had a linear or quadratic relationship (Table S1). To account for multiple tests, we both applied a False Discovery Rate (FDR) correction, and performed a resampling analysis to determine the number of transcripts that would be expected to have a significant relationship by chance alone. We retained the `r nrow(pre.signif.transcripts) - quantile(D$num_signif, probs = 0.95)` (`r round((nrow(pre.signif.transcripts) - quantile(D$num_signif, probs = 0.95)) / nrow(annotation.table), 3) * 100`% of total) transcripts that exceeded the null expectation from the resampling analysis as true positive transcripts for further analyses. Of these transcripts, `r round(sigtable[which(sigtable$Coefficient == "Temperature:Colony"), "Number.significant"] / sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), 2) *100`% (`r prettyNum(sigtable[which(sigtable$Coefficient == "Temperature:Colony"), "Number.significant"], big.mark=",")`) had a non-linear relationship with temperature that would likely have been missed with a standard differential expression experiment (e.g. high vs. low temperature). The proportion of responsive transcripts is similar if we focus only on those transcript with a BLAST hit (`r prettyNum(responsive_transcripts_w_BLAST, big.mark = ",")` significant transcripts out of `r prettyNum(nrow(annotation.table.responsive), big.mark = ",")`, `r round(responsive_transcripts_w_BLAST / nrow(annotation.table.responsive), digits = 3)*100`). However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts [@vijay2013] so this estimate is likely a lower bound for the true proportion of transcripts that are thermally responsive.

@@

In addition, we also examined the specific patterns of shifts from one expression category to another between species. When controlling for overall differences in the number of responsive transcripts, the expression categories of individual transcripts between the two species were not randomly distributed (Stuart-Maxwell test of marginal homogeneity $\chi^2$~4~ = 319, P < 0.001, Fig. S1). Specifically, the two species differed significantly in expression pattern, which captures differences in slope as well as category, for `r prettyNum(sigtable[which(sigtable$Coefficient=="Temperature:Colony"), "Number.significant"], big.mark = ",")` (`r round(sigtable[which(sigtable$Coefficient=="Temperature:Colony"), "Number.significant"] / sigtable[which(sigtable$Coefficient=="Total"), "Number.significant"], 2) * 100`%) of the thermally responsive transcripts.

The *enhanced response* and *tolerance* hypotheses make opposing predictions concerning the overlap in response patterns between the two species (Fig. 2). The *enhanced response hypothesis* posits that temperature adaptation uses existing mechanisms for thermal resistance, which should result in significant overlap in response and fewer transcripts shifting expression categories than expected by chance (Fig. 2, left). In contrast, the *tolerance hypothesis* predicts that transcripts involved in active defense will become non-responsive or shift to other expression categories in the better-adapted species (Fig. 2, right). We tested these predictions by examining if the transcripts upregulated in response to the temperature extreme experienced less frequently by each species (cool temperatures for the warm-climate *A. carolinensis*, and warm temperatures for the cool-climate *A. picea*) displayed the same response profile in the species that more frequently experiences those conditions.

@@

In this work, we have brought reaction norms to the genomic era by characterizing the thermal reactionomes of two temperate ant species, *Aphaenogaster picea* and *A. carolinensis*. At least 2.5% of their transcriptomes are thermally responsive. Our results indicate that these two ant species have different responses to thermal extremes. *A. picea* responds by increasing expression of transcripts related to metabolism, stress response and other protective molecules, whereas *A. carolinensis* decreases expression of transcripts related to metabolism and likely relies on other mechanisms for thermal tolerance. The thermal reactionomes of these two species provide key insights into the genetic basis of thermal tolerance, and a resource for the future study of ecological adaptation in ant species. Finally, the reactionome itself illustrates a new direction for characterizing acclimation and adaptation in a changing climate.

@@

The Trinity transcriptome assembly, FastAnnotator annotation file and Sailfish gene expression quantification files supporting the results of this article are available at the Harvard Forest Archives, HF-113 (http://harvardforest.fas.harvard.edu/data-archive) datasets hf113-38, hf113-41, and hf113-42 (doi:10.6073/pasta/050a6c7c2ff25841b070a0ecab32bc65).

@@

**Figure 1**. Illustration of the patterns against temperature for each of the four expression categories, **Bimodal**, **High**, **Intermediate** and **Low**. The fifth category of **Not Responsive** is not shown as these transcripts would be a line with no clear pattern.

>>>>

---
title: Thermal reactionomes reveal divergent responses to thermal extremes in warm
  and cool-climate ant species
author: ' '
csl: bmcgenomics.csl
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: no
    toc_depth: 2
  html_document:
    keep_md: yes
  word_document: default
includes: header.tex
bibliography: ApTranscriptome.bib
---

@@

We used the reactionome method to identify genes that are thermally responsive in two closely-related eastern North American ant species, *Aphaenogaster carolinensis* and *A. picea* [@umphrey1996; @demarco2015]. *Aphaenogaster* are some of the most common ants in eastern North America [@king2013] where they are keystone seed dispersers [@ness2009; @zelikova2011; @rodriguez-cabal2012]. Ants, and ectotherms in general, have little or no thermal safety margin [@sunday2014] and thus are highly susceptible to climate change [@diamond2012; @kingsolver2013], putting at risk important ecosystem services [@toro2012]. Growth chamber studies have demonstrated that reproduction of *Aphaenogaster* will be compromised by increased tempreatures [@pelini2012], while field studies [@diamond2012] and simulations [@bewick2014] indicate that ant species persistence will depend on combinations of physiology and species interactions. *Aphaenogaster carolinensis* experiences a higher mean annual temperature (MAT) (14.6°C) and less seasonal temperature variation (temperature seasonality = 7,678°) than does *A. picea* (MAT = 4.6°C, seasonality = 10,008°; [@hijmans2005]) at their respective collection sites. In controlled laboratory experiments, these warm- and cold-climate species exhibit corresponding differences in their critical maximum (44.7°C for *A. carolinensis* versus 41.3°C for *A. picea*; see Methods) and minimum temperatures (6.1°C for *A. carolinensis* versus -0.1°C for *A. picea*). These differences between species in their thermal environments and physiological tolerances allowed us to investigate adaptation to both lower and upper thermal extremes in this system.

@@

To summarise, in this project we generated the transcriptomes of two closely-related temperate ant species, and quantified their gene expression across a wide range of thermal conditions. We then evaluated three non-mutually exclusive hypotheses (enhanced response, tolerance and genetic assimilation) of the evolution of thermal adaptation by comparing the number and expression patterns of transcripts between species in response to extreme low and extreme high temperatures. Finally, we used gene ontology information to determine which gene products and pathways are involved in thermal adaptation in the two species.

@@

The combined *Aphaenogaster* transcriptome assembly contained `r prettyNum(nrow(annotation.table), big.mark =",")` transcripts. About half of these (`r prettyNum(nrow(annotation.table.responsive), big.mark =",")`) transcripts had a signficant BLAST hit, of which 50% (25,797) had a top hit to Insecta and 37% (18,854) had a top hit to Formicidae. We performed a BUSCO analysis [@simao2015] to assess the quality of the transcriptome assembly against the arthropod Benchmarking Universal Single-Copy Orthologs (BUSCOs). This analysis revealed that transcriptome is largely complete, as we recovered 1,426 complete single-copy BUSCOs (62%) and an additional 435 fragmented BUSCOs (16%), which is in line with results of Simao et al. [-@simao2015] for transcriptomes of other non-model species. Moreover, only 8% of the BUSCOs were found to be duplicated in the transcriptome, which indicates that the steps (see Methods) we took to collapse homologs in the combined transcriptome of the two species were successful.

We quantified gene expression using the program `Sailfish` [@patro2014], and fitted polynomial regression models to the expression values of each transcript to identify those that had a linear or quadratic relationship (Fig. 1). To account for multiple tests, we both applied a False Discovery Rate (FDR) correction, and performed a resampling analysis to determine the number of transcripts that would be expected to have a significant relationship by chance alone. We retained the `r prettyNum(nrow(pre.signif.transcripts) - quantile(D$num_signif, probs = 0.95), big.mark = ",")` (`r round((nrow(pre.signif.transcripts) - quantile(D$num_signif, probs = 0.95)) / nrow(annotation.table), 3) * 100`% of total) transcripts that exceeded the null expectation from the resampling analysis as true positive transcripts for further analyses (Table S1). Of these transcripts, `r round(sigtable[which(sigtable$Coefficient == "Temperature:Colony"), "Number.significant"] / sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), 2) *100`% (`r prettyNum(sigtable[which(sigtable$Coefficient == "Temperature:Colony"), "Number.significant"], big.mark=",")`) had a non-linear relationship with temperature that would likely have been missed with a standard differential expression experiment (e.g. high vs. low temperature). The proportion of responsive transcripts is similar if we focus only on those transcript with a BLAST hit (`r prettyNum(responsive_transcripts_w_BLAST, big.mark = ",")` significant transcripts out of `r prettyNum(nrow(annotation.table.responsive), big.mark = ",")`, `r round(responsive_transcripts_w_BLAST / nrow(annotation.table.responsive), digits = 3)*100`). However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts [@vijay2013] so this estimate is likely a lower bound for the true proportion of transcripts that are thermally responsive.

@@

In addition, we also examined the specific patterns of shifts from one expression category to another between species. As transcripts may change expression between species due to neutral drift alone, we used the Stuart-Maxwell test of marginal homogeneity to test if the number of responsive transcripts in each expression category differed between the species when controlling for overall differences in the number of responsive transcripts. We found that the expression categories of individual transcripts between the two species were not randomly distributed (Stuart-Maxwell test of marginal homogeneity $\chi^2$~4~ = 319, P < 0.001, Fig. S1). Specifically, the two species differed significantly in expression pattern, which captures differences in slope as well as category, for `r prettyNum(sigtable[which(sigtable$Coefficient=="Temperature:Colony"), "Number.significant"], big.mark = ",")` (`r round(sigtable[which(sigtable$Coefficient=="Temperature:Colony"), "Number.significant"] / sigtable[which(sigtable$Coefficient=="Total"), "Number.significant"], 2) * 100`%) of the thermally responsive transcripts.

The *enhanced response* and *tolerance* hypotheses make opposing predictions concerning the overlap in response patterns between the two species (Fig. 2). The *enhanced response hypothesis* posits that temperature adaptation uses existing mechanisms for thermal resistance, which should result in significant overlap in response and fewer transcripts shifting expression categories than expected by chance (Fig. 2, left). In contrast, the *tolerance hypothesis* predicts that transcripts involved in active defense will become non-responsive or shift to other expression categories in the better-adapted species (Fig. 2, right). We tested these predictions by examining if the transcripts upregulated in response to the temperature extreme experienced less frequently by a species (cool temperatures for the warm-climate *A. carolinensis*, and warm temperatures for the cool-climate *A. picea*) displayed the same response profile in the other species that more frequently experiences those conditions.

@@

In this work, we have brought reaction norms to the genomic era by characterizing the thermal reactionomes of two temperate ant species, *Aphaenogaster picea* and *A. carolinensis*. At least 2% of their transcriptomes are thermally responsive. Our results indicate that these two ant species have different responses to thermal extremes. *A. picea* responds by increasing expression of transcripts related to metabolism, stress response and other protective molecules, whereas *A. carolinensis* decreases expression of transcripts related to metabolism and likely relies on other mechanisms for thermal tolerance. The thermal reactionomes of these two species provide key insights into the genetic basis of thermal tolerance, and a resource for the future study of ecological adaptation in ant species. Finally, the reactionome itself illustrates a new direction for characterizing acclimation and adaptation in a changing climate.

@@

The Trinity transcriptome assembly, FastAnnotator annotation file and Sailfish gene expression quantification files supporting the results of this article are available from the LTER data portal, datasets hf113-38, hf113-41, and hf113-42 (http://dx.doi.org/10.6073/pasta/05ea6464df30efa2f1e2c7439366bf47).

@@

**Figure 1**. Illustration of the patterns against temperature for each of the four expression categories, **Bimodal**, **High**, **Intermediate** and **Low**. The fifth category of **Not Responsive** is not shown.

### 2016-01-15 doc/BMCGenomics_cover_letter.tex
revisions for BMC Genomics

\noindent{Dear Dr. Riesberg,}

We are excited to submit our manuscript \emph{Thermal
reactionomes reveal adaptive responses to thermal extremes in warm and
cold-climate ant species} for consideration as an Original Article in
\emph{Molecular Ecology}. There are
several aspects of the work that we think make it worthy of publication
in \emph{Molecular Ecology}.

@@

In this study, we resolve this problem by extending the reaction
norm approach to the study of gene expression. We define the
\emph{reactionome} as a characterization of the reaction norm for all
genes in an organism's transcriptome across an environmental gradient. 
We believe that this approach is an important contribution as it allows us
to make quantitative comparisons of plasticity in gene expression between
species. Importantly, this analysis does not depend on the functional 
annotation of the genes, and thus is robust to criticisms of "storytelling" 
in RNAseq studies based on annotation of differentially expressed 
genes (Pavlidis et al. 2012 MBE).

@@

\textbf{Suggested reviewers:}

\begin{itemize}
  \item Dr. Nicholas Teets - expertise with insect genomics and transcriptomics
  \item Dr. Vanessa Kellerman - expertise with insect population and quantitative genetics
  \item Dr. Pavlos Pavlidis - critical insight on function annotation and gene ontologies
  \item Dr. Joel Kingsolver - expertise in insect thermal limits
  \item Dr. Daniel Barshis - expertise in ecological genomics
\end{itemize}

\textbf{Suggested editors:}

\begin{itemize}
  \item Dr. Ary Hoffman - expertise with insect population and quantitative genetics
  \item Dr. Stephen Palumbi - expertise in ecological genomics
  \item Dr. Aurelie Bonin - expertise in population genomics
\end{itemize}

>>>>

\noindent{Dear Editor,}

We are writing to submit our manuscript \emph{Thermal
reactionomes reveal divergent responses to thermal extremes in warm and
cold-climate ant species} for consideration as a Research Article. There are
several aspects of the work that we think make it worthy of publication
in \emph{BMC Genomics}.

@@

In this study, we address this problem by extending the reaction
norm approach to the study of gene expression. We define the
\emph{reactionome} as a characterization of the reaction norm for all
genes in an organism's transcriptome across an environmental gradient. 
We believe that this approach is an important contribution as it allows us
to make quantitative comparisons of plasticity in gene expression between
species. Importantly, this analysis does not depend on the functional 
annotation of the genes, and thus is robust to criticisms of "storytelling" 
in RNAseq studies based on annotation of differentially expressed 
genes (Pavlidis et al. 2012 MBE).

@@
